---
title: "Using CVPIA Data Packages to Generate Model Inputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{'Using CVPIA Data Packages to Generate Model Inputs'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
The following vignette provides a method for generating spawning, fry, juvenile, 
and floodplain rearing habitat inputs for the CVPIA salmon life cycle model.
Readers will get a sense of how to use the data and functions from the
[cvpiaFlow](https://flowwest.github.io/cvpiaFlow/) and [cvpiaHabitat](https://flowwest.github.io/cvpiaHabitat/) 
packages. They will also be able to generate additional habitat inputs by modifying the functions 
within this document to accommodate new scenarios for the model.


## Helper functions and Libraries

The following packages are used by the functions. The packages [dplyr](http://dplyr.tidyverse.org/), [tidyr](http://tidyr.tidyverse.org/), and [lubridate](http://lubridate.tidyverse.org/) 
are useful for data manipulations. The package [purrr](http://purrr.tidyverse.org/) is a
functional programming toolkit to replace for loops. 
```{r, warning = FALSE, message = FALSE}
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(cvpiaHabitat)
library(cvpiaFlow)

```

These functions are called by the other habitat setting functions.
```{r}
# returns flow for each month of a watershed during simulation window
get_flow <- function(watershed, years=c(1980, 1999)) {
  
    # get the flow values at the dates
    dplyr::pull(dplyr::filter(dplyr::select(cvpiaFlow::flows_cfs, date, watershed), 
                                             lubridate::year(date) >= years[1], 
                                             lubridate::year(date) <= years[2]), 2)
}

# transforms to array data structure for SIT model input, [watersheds, months, years]
create_SIT_array <- function(input) {

  output <- array(NA, dim = c(nrow(input), 12, ncol(input) / 12))
  index <-  1
  for (i in seq(1, ncol(input), 12)) {
    output[ , , index] <- as.matrix(input[ , i:(i + 11)])
    index <- index + 1
  }
  return(output)

}
```

## Spawning Habitat

The following function outputs the three dimensional array that is used by the
life cycle model to represent spawning habitat for each month and year in the
simulation window. It handles the special case of Upper Sacramento River, which
has two separate WUA area estimates for when the A.C.I.D. boards are in and out. The function
takes the additional argument `month` for Upper Sacramento River in order to use
the correct WUA relationship given the state of the boards. The A.C.I.D boards 
are in April 1st - October 31st. The function also sets the spawning area value
to `NA` for watersheds without spawning. Additionally, here is the source code for [set_spawning_habitat.](https://github.com/FlowWest/cvpiaHabitat/blob/master/R/set-spawning-habitat.R)
```{r}
get_spawn_hab_all <- function(species) {
  
  watersheds <- cvpiaHabitat::modeling_exist %>% 
    dplyr::filter(!is.na(FR_spawn), Watershed != 'Upper Sacramento River', Watershed != 'Upper Mid Sac Region') %>% 
    dplyr::pull(Watershed)
  
  most <- purrr::map_df(watersheds, function(watershed) {
    flows <- get_flow(watershed, years=c(1979, 1999))
    habitat <- cvpiaHabitat::set_spawning_habitat(watershed, 
                                                  species = species,
                                                  flow = flows)
    tibble(
      year = rep(1979:1999, each = 12),
      month = rep(1:12, 21),
      watershed = watershed, 
      hab_sq_m = habitat)
  })
  
  # deal with sacramento special cases
  # upper sac
  up_sac_flows <- get_flow('Upper Sacramento River', years=c(1979, 1999))
  months <- rep(1:12, 21)
  up_sac_hab <- purrr::map2_dbl(months, up_sac_flows, function(month, flow) {
    cvpiaHabitat::set_spawning_habitat('Upper Sacramento River', 
                                       species = species, 
                                       flow = flow, month = month)
  })
  
  up_sac <- tibble(
    year = rep(1979:1999, each = 12),
    month = rep(1:12, 21),
    watershed = 'Upper Sacramento River', 
    hab_sq_m = up_sac_hab)
  
  hab <-   dplyr::bind_rows(most, up_sac) %>% 
    tidyr::spread(watershed, hab_sq_m) %>% 
    dplyr::bind_cols(tibble(`Sutter Bypass` = rep(NA, 252),
                     `Yolo Bypass` = rep(NA, 252),
                     `Upper-mid Sacramento River` = rep(NA, 252),
                     `Lower-mid Sacramento River` = rep(NA, 252),
                     `Lower Sacramento River` = rep(NA, 252),
                     `San Joaquin River` = rep(NA, 252))) %>% 
    tidyr::gather(watershed, habitat, -year, -month) %>% 
    dplyr::mutate(date = lubridate::ymd(paste(year, month, 1, '-'))) %>% 
    dplyr::select(date, watershed, habitat) %>% 
    tidyr::spread(date, habitat) %>% 
    dplyr::left_join(cvpiaData::watershed_ordering) %>% 
    dplyr::arrange(order) %>% 
    dplyr::select(-watershed, -order) %>% 
    create_SIT_array()
  
  return(hab)
}
```

To create habitat arrays for the different species, you can call the above function
with the appropriate species argument.
```{r, eval = FALSE}
fr_spawn <- get_spawn_hab_all('fr') #fall run
sr_spawn <- get_spawn_hab_all('sr') #spring run
st_spawn <- get_spawn_hab_all('st') #steelhead
```

If you wish to modify the above function to make a new scenario, a possible way 
to override the default data output from `cvpiaData::load_baseline_data` is as follows:
```{r, eval = FALSE}
all_model_inputs <- cvpiaData::load_baseline_data('fall')
# user must define get_spawn_hab_all_modified to their specifications
all_model_inputs$IChab.spawn <- get_spawn_hab_all_modified('fr') 
```

Or, you could swap out the watershed's values with a vector of new values. This
vector must be of length 252, with the first value representing the habitat value
for 01/1979 and the last 12/1999.
```{r, eval = FALSE}
fr_modified <- cvpiaData::fr_spawn #copy baseline fall run spawning habitat
new_upper_sac_vals <- 1:252
# upper sacramento is the first watershed, see cvpiaData::watershed_ordering
fr_modified[1,,] <- new_upper_sac_vals 
all_model_inputs$IChab.spawn <- fr_modified[1,,]
```


## Rearing Habitat
This function generates either the fry or juvenile rearing habitat in
each watershed for a specified species. The function handles the two special cases,
the Upper Sacramento River and the Lower-mid Sacramento River. The `set_instream_habitat`
function takes the additional argument `month` for Upper Sacramento River in order to use
the correct WUA relationship given the state of the boards. The A.C.I.D boards 
are in April 1st - October 31st. The Lower-mid Sacramento River has two flow representations,
one above Fremont Weir and one below. For the Lower-Mid Sacramento river, the `set_instream_habitat`
function takes the additional argument `flow2` and calculates the habitat at each 
flow then sums them proportional to the length of stream above and below the weir.
```{r}
get_rear_hab_all <- function(species, life_stage) {

  watersheds <- cvpiaData::watershed_ordering %>% 
    dplyr::filter(!(watershed  %in% c('Sutter Bypass',
                             'Lower-mid Sacramento River', 'Yolo Bypass'))) %>%
    dplyr::pull(watershed)
  
  most <- map_df(watersheds, function(watershed) {
    flows <- get_flow(watershed)
    habitat <- cvpiaHabitat::set_instream_habitat(watershed, 
                                                  species = species,
                                                  life_stage = life_stage, 
                                                  flow = flows)
    tibble(
      year = rep(1980:1999, each = 12),
      month = rep(1:12, 20),
      watershed = watershed, 
      hab_sq_m = habitat)
  })
  
  # deal with sacramento special cases
  # lower-mid sac
  low_mid_sac_flow1 <- get_flow('Lower-mid Sacramento River1')
  low_mid_sac_flow2 <- get_flow('Lower-mid Sacramento River2')
  
  low_mid_sac_hab <- map2_dbl(low_mid_sac_flow1, low_mid_sac_flow2, function(flow, flow2) {
    cvpiaHabitat::set_instream_habitat('Lower-mid Sacramento River', 
                                       species = species, 
                                       life_stage = life_stage, 
                                       flow = flow, flow2 = flow2)
  })
  
  low_mid_sac <- tibble(
    year = rep(1980:1999, each = 12),
    month = rep(1:12, 20),
    watershed = 'Lower-mid Sacramento River', 
    hab_sq_m = low_mid_sac_hab)
  
  hab <- bind_rows(most, low_mid_sac) %>% 
    spread(watershed, hab_sq_m) %>% 
    bind_cols(tibble(`Sutter Bypass` = rep(NA, 240),
                     `Yolo Bypass` = rep(NA, 240))) %>% 
    gather(watershed, habitat, -year, -month) %>% 
    mutate(date = lubridate::ymd(paste(year, month, 1, '-'))) %>% 
    select(date, watershed, habitat) %>% 
    spread(date, habitat) %>% 
    left_join(cvpiaData::watershed_ordering) %>% 
    arrange(order) %>% 
    select(-watershed, -order) %>% 
    create_SIT_array()
  
  return(hab)
}

```

To create habitat arrays for the different species, you can call the above function
with the appropriate species argument.
```{r, eval = FALSE}
fr_fry <- get_rear_hab_all('fr', 'fry') #fall run
sr_fry <- get_rear_hab_all('sr', 'fry') #spring run
st_fry <- get_rear_hab_all('st', 'fry') #steelhead

fr_juv <- get_rear_hab_all('fr', 'juv') #fall run
sr_juv <- get_rear_hab_all('sr', 'juv') #spring run
st_juv <- get_rear_hab_all('st', 'juv') #steelhead

```

If you wish to modify the above function to make a new scenario, a possible way 
to override the default data output from `cvpiaData::load_baseline_data` is as follows:
```{r, eval = FALSE}
all_model_inputs <- cvpiaData::load_baseline_data('fall')
# user must define get_rear_hab_all_modified to their specifications
all_model_inputs$IChab.fry <- get_rear_hab_all_modified('fr', 'fry')
all_model_inputs$IChab.juv <- get_rear_hab_all_modified('fr', 'juv')
```

Or, you could swap out the watershed's values with a vector of new values. This
vector must be of length 240, with the first value representing the habitat value
for 01/1980 and the last 12/1999.
```{r, eval = FALSE}
fr_modified <- cvpiaData::fr_fry #copy baseline fall run fry rearing habitat
new_upper_sac_vals <- 1:240
# upper sacramento is the first watershed, see cvpiaData::watershed_ordering
fr_modified[1,,] <- new_upper_sac_vals 
all_model_inputs$IChab.fry <- fr_modified[1,,]
```

## Floodplain Habitat

This function sets the floodplain rearing habitat for all the watersheds. 
The Lower-mid Sacramento River has two flow representations, one above Fremont 
Weir and one below. The flow value below the weir is used to calculate floodplain
habitat area.
``` {r}
get_floodplain_hab_all <- function(watersheds, species) {

  watersheds_fp <- cvpiaData::watershed_ordering %>% 
    dplyr::filter(!(watershed  %in% c('Sutter Bypass','Yolo Bypass', 
    'Lower-mid Sacramento River', 'Upper Sacramento River',
    'Upper-mid Sacramento River', 'Lower Sacramento River'))) %>%
    dplyr::pull(watershed)
  
  most <- map_df(watersheds, function(watershed) {
    flows <- get_flow(watershed)
    habitat <- cvpiaHabitat::acres_to_square_meters(
      cvpiaHabitat::set_floodplain_habitat(watershed, species, flows))
    
    tibble(
      year = rep(1980:1999, each = 12),
      month = rep(1:12, 20),
      watershed = watershed, 
      hab_sq_m = habitat)
  })
  
  # deal with sac, already in square meters
  # upper sac
  up_sac_flow <- get_flow('Upper Sacramento River')
  up_mid_sac_flow <- get_flow('Upper-mid Sacramento River')
  low_sac_flow <- get_flow('Lower Sacramento River')
  
  up_sac_fp <- cvpiaHabitat::set_floodplain_habitat('Upper Sacramento River', species, up_sac_flow)
  up_mid_sac_fp <- cvpiaHabitat::set_floodplain_habitat('Upper-mid Sacramento River', species, up_mid_sac_flow)
  low_sac_fp <- cvpiaHabitat::set_floodplain_habitat('Lower Sacramento River', species, low_sac_flow)
  
  # lower-mid sacramento 
  low_mid_sac_flows1 <- get_flow("Lower-mid Sacramento River1") 
  low_mid_sac_flows2 <- get_flow("Lower-mid Sacramento River2") 
  low_mid_sac_fp <- cvpiaHabitat::set_floodplain_habitat('Lower-mid Sacramento River', species,
                                                         low_mid_sac_flows1, flow2 = low_mid_sac_flows2)
  sac <- tibble(
    year = rep(rep(1980:1999, each = 12), times = 4),
    month = rep(1:12, 80),
    watershed = rep(c('Upper Sacramento River', 'Upper-mid Sacramento River', 
                      'Lower-mid Sacramento River', 'Lower Sacramento River'), each = 240), 
    hab_sq_m = c(up_sac_fp, up_mid_sac_fp, low_mid_sac_fp, low_sac_fp))
  
  hab <- bind_rows(most, sac) %>% 
    spread(watershed, hab_sq_m) %>% 
    bind_cols(tibble(`Sutter Bypass` = rep(NA, 240),
                     `Yolo Bypass` = rep(NA, 240))) %>% 
    gather(watershed, habitat, -year, -month) %>% 
    mutate(date = lubridate::ymd(paste(year, month, 1, '-'))) %>% 
    select(date, watershed, habitat) %>% 
    spread(date, habitat) %>% 
    left_join(cvpiaData::watershed_ordering) %>% 
    arrange(order) %>% 
    select(-watershed, -order) %>% 
    create_SIT_array()
  
  return(hab)
  
}

```

To create habitat arrays for the different species, you can call the above function
with the appropriate species argument.
```{r, eval = FALSE}
fr_fp <- get_floodplain_hab_all('fr') #fall run
sr_fp <- get_floodplain_hab_all('sr') #spring run
st_fp <- get_floodplain_hab_all('st') #steelhead

```
NOTE: The floodplain area is total wetted area and needs a suitability factor applied
for all watersheds except the Sacramento Reaches, the Sutter and Yolo Bypasses, and the
North and South Deltas which the floodplain area has already had a suitability criteria
applied. See the apply suitability [function](https://flowwest.github.io/cvpiaHabitat/reference/apply_suitability.html) for more details.

If you wish to modify the above function to make a new scenario, a possible way 
to override the default data output from `cvpiaData::load_baseline_data` is as follows:
```{r, eval = FALSE}
all_model_inputs <- cvpiaData::load_baseline_data('fall')
# user must define get_floodplain_hab_all_modified to their specifications
all_model_inputs$floodP <- get_floodplain_hab_all_modified('fr')
```

Or, you could swap out the watershed's values with a vector of new values. This
vector must be of length 240, with the first value representing the habitat value
for 01/1980 and the last 12/1999.
```{r, eval = FALSE}
fr_modified <- cvpiaData::fr_fp #copy baseline fall run floodplain rearing habitat
new_upper_sac_vals <- 1:240
# upper sacramento is the first watershed, see cvpiaData::watershed_ordering
fr_modified[1,,] <- new_upper_sac_vals 
all_model_inputs$floodP <- fr_modified[1,,]
```

